
\preface[Acknowledgments]

Research is not an individual effort, but requires the professional
and personal support of many people. It is impossible to list
everybody here, so thanks to all of you. Thanks for supporting my
work, for that talk during the coffee break, for that assistance
with the English, for that pat on the back, for those breakfasts
early in the morning, for those ``sport moments'', for those daily
\emph{tupper} meals, for being the lighthouse in the night\dots
Thanks for everything.

To Inma, for turning difficult times into a smile. This work is as
much yours as it is mine.

To my family for making me feel special, particularly to my parents
since all I am is due to them.

To Juan Carlos Trujillo, for all the shared moments and for his
decisive and crucial encouragement to go on.

To Jes{\'u}s Pardillo and Octavio Glorio, for their friendship, for
all the discussions, for their support and for their involvement in
the implementation of this research.

To Emilio Soler, during these years of ``Cuban connection'' and
during this last sprint, working side by side.

To Luisa Micó and Jose Manuel I\~{n}esta, for allowing me to start
working in the Department of Software and Computing Systems of the
University of Alicante some years ago.

To my colleagues in the Lucentia research group in the Department of
Software and Computing Systems of the University of Alicante,
Cristina Cachero, Jose Jacobo Zubcoff, Sergio Luján, Rafael Romero
and Lily Muñoz for their assistance, fruitful discussions and
support.

To all the members of the IWAD research group in the Department of
Software and Computing Systems of the University of Alicante, and
particularly to Jaime Gómez, Irene Garrigós and Santiago Meliá for
their continued support.

To all the members of the GPLSI research group in the Department of
Software and Computing Systems of the University of Alicante, for
all the shared moments.

I would also like to express my deepest gratitude to Prof. Dr.
Gottfried Vossen and Dr. Jens Lechtenbörger for hosting my stay at
the University of Münster during the summer of 2007. Their interest
in my work, valuable discussions and encouragement had a great
influence on the progress of my research.

Finally, I wish to thank Prof. Dr. Eric Yu and Dr. Jordi Cabot for
their support during my research period in the University of Toronto
in the summer of 2008.
