%%%%%%%%%%%%%%%%%%%%%% pref.tex %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample preface
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%%%%%%%%%% Springer-Verlag %%%%%%%%%%%%%%%%%%%%%%%%%%

\preface

%The engineering methodologies are an essential component for the Web applications development process, this is because through by a series of well-defined methods it is possible to undertake a systematic and structured development process. 

%In recent years, there have been new proposals to address the development of Web applications, some of them are mainly focused in the representation of the Web application at some level of abstraction (conceptual model), others meanwhile, are focused on specific tasks of the development process leaving aside the requirements phase. Moreover, because of the increasing complexity of the Web applications (i.e., changes in the platform implementation technology) and the multiple audiences involved in their use (i.e., heterogeneous audience), the requirements phase is more difficult to perform and to maintain. As a result, a problem emerges in these proposals: the absence of a design guide that facilitates the development of the Web applications based on the user's needs and expectations. 

En los últimos años, han surgido nuevas propuestas para abordar el desarrollo de aplicaciones Web, algunas de ellas se centran principalmente en la representación de la aplicación Web en un cierto nivel de abstracción (modelo conceptual), otros por su parte, se centran en tareas específicas de la proceso de desarrollo dejando de lado la fase de requisitos. Además, debido a la creciente complejidad de las aplicaciones Web (es decir, los cambios en la tecnología de implementación) y las múltiples audiencias implicadas en su uso (audiencia heterogénea), la fase de requisitos es más difícil de realizar y mantener. Como resultado, un problema surge en estas propuestas: la ausencia de una guía de diseño que facilite el desarrollo de las aplicaciones Web basada en las necesidades del usuario y las expectativas.

%To overcome the lack of such a process, this PhD Thesis proposes a contribution to a existing model-driven approach for the development of Web 1.0 applications. Specifically, we propose the requirements specification in a conceptual model based in the \emph{i*} goal-oriented modeling framework, with which, the automatic derivation of the Web application conceptual models are possible. A requirements managing support is also proposed to avoid problems at the conceptual level with regard to (i) requirements traceability, (ii) change impact analysis and (iii) the design choices based on non-functional requirements maximization. Finally, as a proof of concept a set of \emph{Eclipse} plugin's have been implemented. These are applied in a case study.

Para superar la falta de dicho proceso, la tesis doctoral presenta una aproximación dirigida por modelos para el desarrollo de aplicaciones Web. En concreto, se propone la especificación de los requisitos de la aplicación Web un modelo conceptual basado en el marco de modelado orientado a objetivos \emph{i*}. Al realizarse la especificación de los requisitos en un modelo, la derivación automática de los modelos conceptuales que constituyen una aplicaciones Web es posible. Además, se presenta el soporte para la gestión de requisitos con lo cual será factible evitar problemas a nivel conceptual con (i) trazabilidad de requisitos, (ii) análisis de impacto y (iii) las decisiones de diseño que esten basadas en la maximización de los requisitos no-funcionales. Por último, como una prueba de concepto se ha implementado la propuesta en la plataforma \emph {Eclipse} y se demuestra su aplicabilidad en un caso de estudio.

%This PhD Thesis is composed of a set of published and submitted papers. In order to write this PhD Thesis as a collection of papers, several requirements must be taken into account as stated by the University of Alicante. With regard to the content of the PhD Thesis, it must specifically include a summary with regard to the description of the initial hypotheses, research objectives, and the collection of publications itself. This summary of the PhD Thesis includes the research results and the final conclusions. Finally, this summary corresponds to the part~\ref{p1} of this PhD Thesis (chapter~\ref{c1} has been written in Spanish while chapter~\ref{c2} is in English).

La tesis doctoral se compone de un conjunto de artículos publicados y algunos en proceso de revisión. Para escribir la tesis como una colección de artículos se deben de tener en cuenta una serie de requisitos según lo estipulado por la Universidad de Alicante. Con respecto al contenido, la tesis doctoral debe incluir un resumen, las hipótesis iniciales, los objetivos de la investigación y la colección de publicaciones. Finalmente, el resumen corresponde con la parte \ref{p1} de la tesis doctoral.

%It is important to highlight that this PhD Thesis has been developed under the PhD program \emph{``Aplicaciones de la Informática''} of the Department of Software and Computing Systems (\emph{Departamento de Lenguajes y Sistemas Informáticos}, DLSI) of the University of Alicante. This PhD work was funded by the CONACYT (\emph{Consejo Nacional de Ciencia y Tecnología, Mexico} and University of Sinaloa, Mexico.

Es importante destacar que la tesis doctoral se ha desarrollado dentro del programa de doctorado \emph{``Aplicaciones de la Informática''} del Departamento de Lenguajes y Sistemas Informáticos (DLSI) de la Universidad de Alicante. El trabajo doctoral ha sido financiado por el CONACYT (\emph{Consejo Nacional de Ciencia y Tecnología, México} y la Universidad Autónoma de Sinaloa, México.

%Finally, this research was developed under the following projects: MANTRA (GRE09-17) from the University of Alicante, SERENIDAD (PEII-11-0327-7035) from Junta de Comunidades de Castilla La Mancha (Spain) and by the MESOLAP (TIN2010-14860) project from the Spanish Ministry of Education and Science and QUASIMODO (PAC08-0157-0668)
%projects from the Castilla-La Mancha Ministry of Education and Science (Spain).

Finalmente, la investigación se desarrolló bajo los siguientes proyectos: Mantra (GRE09-17) de la Universidad de Alicante, SERENIDAD (PEII-11-0327-7035) de la Junta de Comunidades de Castilla La Mancha (España), MESOLAP (TIN2010 -14.860) del  Ministerio Español de Educación y Ciencia y por el proyecto Quasimodo (PAC08-0157-0668) del Ministerio de Educación y Ciencia de Castilla-La Mancha (España).

%% Please "sign" your preface
\vspace{1cm}
\begin{flushright}\noindent
Alicante, Julio 2011\hfill {\it Jos{\'e} Alfonso Aguilar Calder{\'o}n}\\
\end{flushright}
